\section{NLP Tools and Libraries}

Numerous tools and libraries facilitate the development and implementation of different Natural Language Processing tasks.

\subsection{Python Libraries}

\subsubsection{Natural Language Toolkit (NLTK)}

\textbf{NLTK} is one of the most widely used tools for implementing NLP applications in Python. It is:
\begin{itemize}
    \item Available for Windows, Mac OS X, and Linux
    \item Open source (Apache License Version 2.0)
    \item Primarily developed for English NLP, but flexible enough for other languages like Spanish
\end{itemize}

\textbf{Features:}
\begin{itemize}
    \item Interfaces to more than fifty corpora and lexical resources (e.g., WordNet)
    \item Libraries for classification, tokenization, stemming, morphosyntactic tagging, syntactic parsing, and semantic analysis
    \item Can train morphosyntactic taggers in Spanish from tagged corpora
    \item Allows integration of external taggers for Spanish
\end{itemize}

\subsubsection{spaCy}

\textbf{spaCy} is a Python library widely used in both academic and industrial settings.

\textbf{Features:}
\begin{itemize}
    \item Incorporates various text normalization pipelines (tokenization, lemmatization, etc.)
    \item Supports more than sixty languages
    \item Includes recent neural models for NLP tasks (e.g., Named Entity Recognition)
    \item Includes pre-trained models like BERT
    \item Open source for commercial use (MIT License)
\end{itemize}

\subsubsection{Gensim (Generate Similar)}

\textbf{Gensim} is a Python library for working with NLP.

\textbf{Features:}
\begin{itemize}
    \item Topic modeling
    \item Document indexing
    \item Document search using techniques like vector embeddings
\end{itemize}

\subsection{Java Libraries}

\subsubsection{LingPipe}

\textbf{LingPipe} is a Java library for developing NLP applications.

\subsubsection{Apache OpenNLP}

\textbf{Apache OpenNLP} is a Java library for NLP tasks, providing tools for various text processing operations.

\subsubsection{Stanford NLP}

\textbf{Stanford NLP} is a suite of Java libraries for NLP, offering various tools for text analysis and processing.

\subsubsection{Stanford CoreNLP}

\textbf{Stanford CoreNLP} is a comprehensive Java library for NLP tasks.

\textbf{Features:}
\begin{itemize}
    \item Incorporates models for NLP tasks in Spanish
    \item Handles Spanish-specific characteristics:
    \begin{itemize}
        \item Separation of word contractions (e.g., ``del'' = ``de'' + ``el'') during tokenization
        \item Identification of verb tenses and moods (e.g., present indicative) in morphosyntactic tagging
    \end{itemize}
    \item Implemented in Java but provides command-line access, enabling integration with Python
    \item Can be used with NLTK in Python code for Spanish morphosyntactic tagging
\end{itemize}

\subsection{Node.js Libraries}

\subsubsection{Natural}

\textbf{Natural} is a library for implementing NLP tasks in Node.js, providing various text processing capabilities for JavaScript applications.

